Functional timeseries reveal no systemic shifts for vertebrate communities

### 2021-04-19 paper/manuscript.Rmd
Discussion, list of papers to go back to

---
title: "Functional trait data tell new story about biodiversity change"
author: 
  - name: "Kari E A Norman"
    affiliation: ucb
  - name: "Carl Boettiger"
    affiliation: ucb, ropensci
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: ropensci
    address: "The rOpenSci Project, University of California Berkeley, Berkeley CA 94720-3114, USA" 
abstract: |

journal: "Ecography"
date: "`r Sys.Date()`"
bibliography: refs.bib
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: word_document
  #rticles::elsevier_article:
   # includes:
    #  in_header: preamble.tex
editor_options: 
  markdown: 
    wrap: 72
---

@@

Ecological communities are experiencing unprecedented change as a result
of anthropogenic pressures such as climate change, land use change, and
invasive species. Impacts of these pressures are well documented at a
global scale by an accelerating global extinction rate [@barnosky2011],
and fundamental changes in some of the most well-studied systems (e.g.
coral bleaching). At the local scale however, species diversity tells a
different story. Recent synthesis of local trends in biodiversity over
time have found no net change in local biodiversity despite ongoing
species turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017]
and evidence of a decoupling of community composition from richness
[[@brose2016]; [@gotelli2017]. Potential hypothesis for the balance of
etc..<!-- potential indication of biological homogenization due to shifting ranges and/or invasive species-->
While communities are clearly changing, our most common species-based
approaches do not fully capture the nature of that change.

@@

Functional diversity also has the potential to shed light on critical
ecosystem and community processes potentially impacted by anthropogenic
change. For example, there is increasing evidence that functional
diversity is a better predictor of ecosystem function than species-based
metrics [@cadotte2011; @gagic2015], and that different facets of
functional diversity also play a role in maintaining ecosystem stability
[@craven2018; @morin2014]. Further, biogeographic assessments of
functional diversity may give foundational insight into how community
assembly plays out across systems [@violle2014; @weiher2011].

@@

Assessments of broad-scale temporal change in functional diversity have
previously been limited by a lack of functional trait data. The majority
of work has therefore focused largely on system-specific studies with
traits collected *in situ*. Ongoing efforts to assemble functional
traits for a variety of taxa have made synthesis of existing community
assemblage data and functional traits possible for the first time,
providing initial looks at how functional diversity changes on a
broad-scale for specific taxa (e.g. fish[@trindade-santos2020], birds
[@jarzyna2016; @barnagaud2017]).
<!-- so what do these studies say? does that go here? --> However, to
date there has been no cross-taxa assessment of functional change
through time across a broad geographic and taxonomic extent.

Here we perform the first multitaxa, multisystem assessment of
functional diversity change through time. We assess thousands of mammal,
bird, and amphibian functional diversity timeseries to 1) determine how
functional diversity change compares to species change and 2) identify
three distinct types of functional change common across communities.
While our data represent the largest collection of functional diversity
timeseries to date, they are not a representative sample of systems
experiencing biodiversity change. We do not therefore seek to classify
the global trend in functional diversity change, but rather assess the
utility of a functional lens in understanding biodiversity change.

@@

Functional diversity metrics were calculated using the dbFD function
from the FD R package. Here we report functional richness (*FRic*),
functional evenness (FEve), and functional diveregence (FDiv) which
together describe three complementary characteristics of the functional
space [@mason2005]. FRic assesses the volume of the trait space occupied
by species in the community. FEve describes how species are distributed
across the trait space and how abundance is distributed across species.
FDiv measures the degree to which species and their abundances maximize
differences in the functional space. Higher values of FDiv therefore
correspond to communities where many highly abundant species are on the
edges of the trait space.

@@

We corrected for bias in functional diversity metrics due to changes in
species richness by calculating the standardized effect size (SES) for
each metric from null estimates [@swenson2012]. Null model corrections
allow us to assess the degree to which the observed functional diversity
metric deviates from the value expected by chance in a randomly
assembled community. Null estimates were calculated for each rarefied
sample by randomly sampling species from the species pool for each year
and randomly assigning observed abundances to species. Species pools
included all species observed for a timeseries. This process was
repeated 500 times to get an estimate and standard deviation of the null
expectation for the metric for that timeseries. We used these values to
calculate SES using the following formula:
$SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$. Unless otherwise
specified, functional diversity results are reported for standardized
values for the rest of the paper.

@@

We estimated overall trends for each diversity metric using a Linear
Mixed Effects model with a random slope and intercept for each
timeseries nested within study. Differences in trends by taxa, biome,
and realm were also modeled individually as covariates in models of the
same structure.

@@

We found no significant temporal trend in species richness or functional
diversity metrics for either observed or standardized metrics. There
continued to be no trend for studies broken down by taxa, biome or
realm. We did find a significant overall decrease in Jaccard similarity,
indicating increasing turnover through time.

At the study level, the majority of studies (78%) experienced a
significant temporal change in species richness and all functional
diversity metrics. 92% of sites experienced a significant trend for
species richness, 88% for functional functional richness, 96% for
functional divergence, 88% for functional evenness, and 35% for Jaccard
similarity. Only four studies experienced no significant trend in
species richness and were also characterized by significantly decreasing
functional evenness. There was only one significant pairwise correlation
between the metrics, a positive relationship between Functional
Divergence and Functional Evenness, indicating that studies experiencing
an increase in Divergence frequently experience and increase in
Evenness.

Cluster analysis grouped all but a few studies into a cluster
characterized by non-siginficant or very small significant slopes for
all metrics. We identified four other clusters containing six studies
that were characterized by outlier trends in at least one of the metrics
and near-zero trends in the rest. The first contained two studies with
the most strongly decreasing evenness. The second, two studies with the
most strongly increasing functional divergence and increasing functional
richness. The third contained one study with strongly decreasing
functional divergence and richness, and the fourth strongly decreasing
species richness with increasing functional richness.

@@

Our study represents the largest broad-scale multi-taxa assessment of
functional change through time, giving a first look at overall trends in
functional diversity and the range of changes happening in mammal, bird,
and amphibian communities. We found no systematic temporal change in any
measured dimension of functional diversity or species richness, even
within specific realms, biomes, or taxonomic groups. Species-based
results are consistent with other broad scale analyses of biodiversity
change through time [@dornelas2014] that found no overall trend in
species richness despite increasing turnover. While nearly all of the
studies included in our data experienced a significant increase or
decrease in species richness, they were observed in equal numbers
resulting in an overall trend not significantly different from zero.
Surprisingly, there was also no general trend in functional diversity
metrics indicating that, similar to species richness, communities are
experience a wide range of functional shifts. Non significant results
for study characteristics confirm that functional changes cannot be
explained by the taxa, realm, or climate characteristics of a given
study.

Individual studies in our dataset demonstrate the variety of species and
functional shifts that can happen concurrently in a community. Notably,
many studies did not experience increasing turnover (65%) despite having
significant change in functional dimensions.

-   very few sites experienced increasing turnover, but the ones that
    did were significant!

-   what does it me for turnover not to be increasing but the function
    of communities to be changing

@@

@@

>>>>

---
title: "Functional timeseries reveal no systemic shifts for vertebrate communities"
author: 
  - name: "Kari E A Norman"
    affiliation: ucb
  - name: "Carl Boettiger"
    affiliation: ucb, ropensci
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: ropensci
    address: "The rOpenSci Project, University of California Berkeley, Berkeley CA 94720-3114, USA" 
abstract: |

journal: "Ecography"
date: "`r Sys.Date()`"
bibliography: refs.bib
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: #word_document
  rticles::elsevier_article:
  includes:
   in_header: preamble.tex
editor_options: 
  markdown: 
    wrap: 72
---

@@

Ecological communities are experiencing unprecedented change as a result
of anthropogenic pressures such as climate change, land use change, and
invasive species. Impacts of these pressures are well documented at a
global scale by an accelerating global extinction rate [@barnosky2011],
and fundamental changes in some of the most well-studied systems (e.g.
coral bleaching). At the local scale however, species diversity tells a
different story. Recent synthesis of local trends in biodiversity over
time have found no net change in local biodiversity despite ongoing
species turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017]
and evidence of a decoupling of community composition from richness
[@brose2016; @gotelli2017] . Potential hypothesis for the balance of
etc..<!-- potential indication of biological homogenization due to shifting ranges and/or invasive species-->
While communities are clearly changing, our most common species-based
approaches do not fully capture the nature of that change.

@@

Loss of functional diversity is also linked to impacts on critical
ecosystem and community processes. For example, there is increasing
evidence that functional diversity is a better predictor of ecosystem
function than species-based metrics [@cadotte2011; @gagic2015], and that
different facets of functional diversity play essential roles in
maintaining ecosystem stability [@craven2018; @morin2014]. Further,
maintenance of functional diversity is key for maximizing adaptive
potential of communities under climate change.

<!-- Functional loss is often implicit in work linking species loss to impacts on ecosystem processes despite a wide range of documented relationships between functional and species change [@biswas2011]. Examining functional and species change side by side across a variety of communities -->

@@

Assessments of broad-scale temporal change in functional diversity have
previously been limited by a lack of functional trait data. The majority
of work has therefore focused largely on system-specific studies with
traits collected *in situ*. Ongoing efforts to assemble functional
traits for a variety of taxa have made synthesis of existing community
assemblage data and functional traits possible for the first time,
providing initial looks at how functional diversity changes on a
broad-scale for specific taxa (e.g. fish[@trindade-santos2020], birds
[@jarzyna2016; @barnagaud2017]).
<!-- so what do these studies say? does that go here? --> However, to
date there has been no cross-taxa assessment of temporal functional
change for a broad geographic and taxonomic extent.

Here we perform the first multi-taxa, multi-system assessment of
functional diversity change through time. We assess thousands of mammal,
bird, and amphibian functional diversity timeseries to 1) determine how
functional diversity change compares to species change and 2) assess the
types of functional shifts common across communities.
<!-- While our data represent the largest collection of functional diversity timeseries to date, they are not a representative sample of systems experiencing biodiversity change. We do not therefore seek to classify the global trend in functional diversity change, but rather assess the utility of a functional lens in understanding biodiversity change.-->

@@

Functional diversity metrics were calculated using the *dbFD* function
from the *FD* R package. Here we report functional richness (*FRic*),
functional evenness (*FEve*), and functional diveregence (*FDiv*) which
together describe three complementary characteristics of the functional
space [@mason2005]. *FRic* assesses the volume of the trait space
occupied by species in the community. *FEve* describes how species are
distributed across the trait space and how abundance is distributed
across species. *FDiv* measures the degree to which species and their
abundances maximize differences in the functional space. Higher values
of *FDiv* therefore correspond to communities where many highly abundant
species are on the edges of the trait space.

@@

We corrected for bias in functional diversity metrics due to changes in
species richness by calculating the standardized effect size (SES) for
each metric from null estimates [@swenson2012]. Null model corrections
allow us to assess the degree to which the observed functional diversity
metric deviates from the value expected by chance in a randomly
assembled community. Null estimates were calculated for each rarefied
sample by randomly sampling species from the species pool for each year
and randomly assigning observed abundances to species. Species pools
included all species observed for a timeseries. This process was
repeated 500 times to get an estimate and standard deviation of the null
expectation for the metric for that timeseries. We used these values to
calculate SES using the following formula:
$SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$. Unless otherwise
specified, functional diversity results are reported for standardized
values for the remainder of the paper.

@@

We estimated overall trends for each diversity metric using a Linear
Mixed Effects model with a random slope and intercept for each
timeseries nested within study. Differences in trends by taxa, biome,
and realm were each modeled individually as covariates in models of the
same structure.

@@

We found no significant temporal trend in species richness or functional
diversity metrics for either observed or standardized metrics. There was
also no trend for studies broken down by taxa, biome or realm. We did
find a significant overall decrease in Jaccard similarity, indicating
increasing turnover through time.

At the study level, the majority of studies (78%) experienced a
significant temporal change in species richness and all functional
diversity metrics. 92% of sites experienced a significant trend for
species richness, 88% for functional functional richness, 96% for
functional divergence, 88% for functional evenness, and 35% for Jaccard
similarity. Only four studies experienced no significant trend in
species richness and were also characterized by significantly decreasing
functional evenness. There was only one significant pairwise correlation
between the metrics, a positive relationship between Functional
Divergence and Functional Evenness, indicating that studies experiencing
an increase in Divergence frequently experience an increase in Evenness.

Cluster analysis grouped all but a few studies into a cluster
characterized by non-significant or very small significant slopes for
all metrics. We identified four other clusters containing six studies
that were characterized by outlier trends in at least one of the metrics
and near-zero trends in the rest. The first contained two studies with
the most strongly decreasing evenness. The second, two studies with the
most strongly increasing functional divergence and increasing functional
richness. The third contained one study with strongly decreasing
functional divergence and richness, and the fourth strongly decreasing
species richness with increasing functional richness.

@@

Our study represents the largest broad-scale multi-taxa assessment of
functional change through time, giving a first look at overall trends in
functional diversity and the range of changes happening in mammal, bird,
and amphibian communities. Consistent with previous species-based
syntheses, we found no overall trend in species diversity accompanied by
increasing turnover through time [@dornelas2018]. While nearly all of
the studies included in our data experienced a significant increase or
decrease in species richness, they were observed in equal numbers
resulting in an overall trend not significantly different from zero.

Surprisingly, we did not detect an overall trend in any of the observed
or corrected functional diversity metrics. Similar to species richness,
communities are experiencing a wide range of functional shifts that
balance each other out. While our dataset likely under-samples areas
with most direct human impact, anthropogenic change is hypothesized to
impact even the most remote systems. Conservatively, our findings
therefore indicate that in unperturbed mammal, bird, and amphibian
communities there is no evidence of systemic functional richness loss or
functional change. A lack of trend for data broken down by realm, biome,
and taxonomic group gives further evidence that directional functional
change is absent from all systems observed in our dataset.

Individual studies in our dataset demonstrate the variety of species and
functional shifts that can happen concurrently in a community. With only
one significant pairwise relationship between Functional Divergence and
Functional Evenness, all other possible combinations of species and
functional change were observed in at least one study. Notably, there is
no relationship between species richness change and change in functional
richness, meaning that even for studies that may be experiencing species
losses there is not necessarily an accompanying loss of functional
richness and in some cases there may even be an increase. These
communities are gaining new species that are functional distinct enough
to make up for functional losses due to species losses, a pattern that
would be further magnified if the species lost are not functionally
distinct members of the assemblage.

The overall trend of increasing turnover is consistent with work
indicating higher than expected background rates of turnover in
contemporary communities [@gotelli2017; @magurran2018]. However, many
studies did not experience increasing turnover (65%), suggesting that it
is not universal. Even for those studies without a turnover trend,
significant functional shifts occurred. While some communities may be
changing more rapidly than usual, that turnover is not necessarily an
indication of specific kind of functional change.

Our results have important implications for how we approach work linking
species loss to ecosystem processes, which often implicitly assume
concurrent functional loss. There is no evidence of directional
functional loss or even functional shifts accompany species loss in
these natural communities. Building on a rich body of work linking
species diversity changes to changes in ecosystem function, resilience,
and stability (citations), it will be essential to tease apart effects
due to species change from those due to functional shifts if we want to
infer impacts in natural communities.

<!-- The absence of systematic loss of functional richness or functional divergence suggests that widespread biological homogenization is not occurring in sampled communities. Biological homogenization occurs when endemic or rare species are lost and replaced with generalist or non-native species, resulting in a loss of -->

As the first broad scale analysis of temporal functional diversity
change we make x broad conclusions. First,

Above all our study shows that patterns of functional change can not be
inferred from

-   how are these findings consistent with expectations for hypothesized
    shifts in functional change?

@@

-   other unmeasured taxa may be harder hit? different patterns for fish
    assemblages

-   contextualize functional shifts in other temporal studies

@@

Papers to look at:

Magurran et al 2018

Magurran et al 2019 (temporal beta diversity)

McGill 2015

Crossley 2020 - not sure what to do with it

Liang et al 2019 - functional homogenization

Finderup Nielsen et al 2019 - biotic homogenization

Dornelas et al 2019

MacLean et al 2018

Hilldebrand 2018

**Ideas to tie in:**

-   species richness loss linked to impacts on processes, in doing so
    often inferring functional loss, but the two do not go hand in hand
    in natural communities

-   no evidence of biological homogenization

-   how is this consistent with theory
